---
title: "Day 1 – Spatial & Spatio-Temporal Modelling"
author: "Olatunji Johnson"
freeze: auto
format:
  html:
    toc: true
    code-fold: true
editor: visual
---

# Overview

Today we introduce modern geostatistical modelling using inlabru, covering:

1.  Spatial models

    -   Binomial likelihood (prevalence)
    -   Poisson likelihood (case counts)
    -   SPDE spatial random fields
    -   Prediction surfaces
    -   Exceedance probabilities

2.  Spatio-temporal models

    -   Handling time-indexed disease data
    -   RW1/RW2 temporal structure
    -   Spatio-temporal Gaussian fields
    -   Predicting in time and space

These form the foundation for Days 2 and 3.

============================================================

**PART A — Spatial models**

============================================================

## 1. Load data

These data were generated during preprocessing and saved to data/

```{r}
library(tidyverse)
library(INLA)
library(inlabru)
library(ggplot2)
library(sf)
library(stars)
library(rnaturalearth)
library(rnaturalearthdata)

nga_admin1 <- ne_states("Nigeria", returnclass = "sf") %>%
st_transform(32632)

spatial_binom <- read_csv("data/spatial_binomial_data.csv")
spatial_pois  <- read_csv("data/spatial_poisson_data.csv")
```

## 2. Convert to sf for mapping

We need an `sf` layer to overlay boundaries, plot points, and feed coordinates into SPDE meshes.

```{r}
spatial_binom_sf <- spatial_binom %>%
  st_as_sf(coords = c("utm_x", "utm_y"), crs = 32632)
spatial_pois_sf <- spatial_pois %>%
  st_as_sf(coords = c("utm_x", "utm_y"), crs = 32632)
```

## 3. Visualise sampling locations

This helps confirm:

-   UTM conversion is correct
-   Points sit inside Nigeria
-   Prevalence has spatial structure

```{r}
ggplot() +
geom_sf(data = spatial_binom_sf, aes(colour = prevalence_true)) +
geom_sf(data = nga_admin1, fill = NA) +
scale_colour_viridis_c(name="Prevalence") +
theme_minimal()
```

## 4. Build the SPDE mesh

The mesh defines the geometry on which the latent spatial field is approximated.

```{r}
coords <- st_coordinates(spatial_binom_sf)

mesh <- inla.mesh.2d(
loc = coords,
max.edge = c(100000, 200000), # inner and outer triangle sizes
cutoff = 20000,  # merge points closer than 20km
offset = c(50000, 200000)  # extend mesh beyond convex hull
)

plot(mesh)
points(coords, col="red", pch=16)
```

A finer mesh → more accurate, more expensive.\
A coarser mesh → faster, less precise.\
We balance cost vs smoothness.

## 5. SPDE model (PC priors)

Penalised complexity (PC) priors make spatial models robust and interpretable.

```{r}
spde <- inla.spde2.pcmatern(
mesh = mesh,
prior.range = c(300000, 0.5),
prior.sigma = c(1.0, 0.01)
)
```

## 6. Spatial BINOMIAL model

The model is:

$$logit(pi​)=\beta_0​+\beta_1​ elevation_i​+ \beta_2​ndvi_i​+ \beta_3 ​urban_i​+ S(s_i​)$$

Where $S(\mathbf{s})$ is a spatial Gaussian field.

```{r}
components_binom <- ~
Intercept(1, model="linear") +
beta_elev(elevation, model="linear") +
beta_ndvi(ndvi, model="linear") +
beta_urban(urban, model="linear") +
field(geometry, model = spde)

lik_binom <- like(
family="binomial",
data=spatial_binom_sf,
formula = n_pos ~ Intercept + beta_elev + beta_ndvi + beta_urban + field,
Ntrials = n_tested
)

#| cache: true
fit_binom <- bru(components_binom, lik_binom)
summary(fit_binom)
```

## 7. Spatial POISSON model

Often used for malaria case counts or incidence.

Model: $$\log(\lambda_i)= \beta_0 + \beta_1 X_i + S(s_i)$$

```{r}
components_pois <- ~
Intercept(1, model="linear") +
beta_elev(elevation, model="linear") +
beta_ndvi(ndvi, model="linear") +
beta_urban(urban, model="linear") +
field(geometry, model = spde)

lik_pois <- like(
family="poisson",
data=spatial_pois_sf,
formula = cases ~ Intercept + beta_elev + beta_ndvi + beta_urban + field
)

#| cache: true
fit_pois <- bru(components_pois, lik_pois)
summary(fit_pois)

```

## 8. Predictive surfaces (binomial example)

```{r}
pred_grid <- read_csv("data/prediction_grid_utm.csv") %>%
st_as_sf(coords=c("utm_x","utm_y"), crs=32632)

pred_binom <- predict(
fit_binom,
newdata = pred_grid,
formula = ~ plogis(Intercept + beta_elev + beta_ndvi + beta_urban + field),
n.samples = 1000
)
```

Rasterise:

```{r}
pred_binom_r <- st_rasterize(pred_binom["mean"], dx=10000, dy=10000)
pred_binom_r <- pred_binom_r[st_union(nga_admin1)]
```

Plot:

```{r}
ggplot() +
geom_stars(data=pred_binom_r) +
scale_fill_viridis_c(name="Prev") +
geom_sf(data=nga_admin1, fill=NA) +
theme_minimal()
```

## 9. Exceedance probability (binomial)

Goal:

$$P(prevalence>0.5∣data)$$

```{r}
exc_samples <- generate(
fit_binom,
newdata = pred_grid,
formula = ~ plogis(Intercept + beta_elev + beta_ndvi + beta_urban + field),
n.samples = 1000
)

pred_binom$exc_prob <- rowMeans(exc_samples > 0.5)
```

Map:

```{r}
exc_r <- st_rasterize(pred_binom["exc_prob"], dx=10000, dy=10000)
exc_r <- exc_r[st_union(nga_admin1)]

ggplot() +
geom_stars(data=exc_r) +
scale_fill_viridis_c(name="P(prev>0.5)", limits=c(0,1)) +
geom_sf(data=nga_admin1, fill=NA) +
theme_minimal()
```

============================================================

**PART B — Spatio-Temporal Modelling** ============================================================

We now extend:

$$S(s)⟶S(s,t)$$

A common structure:

1.  Spatial effect: SPDE

2.  Temporal effect: random walk (RW1 or RW2)

3.  Interaction: separable $S(s,t)=S_s(s)+S_t(t)$

```{r}
st_binom <- read_csv("data/spatiotemporal_binomial_data.csv") %>%
  st_as_sf(coords=c("utm_x","utm_y"), crs=32632)

head(st_binom)
```

Should contain:

-   time

-   n_tested, n_pos

-   covariates

-   UTM coordinates

## 11. Spatio-temporal components

Use: - RW1 for time - SPDE for space - Additive interaction

```{r}
components_st <- ~
Intercept(1, model="linear") +
beta_elev(elevation, model="linear") +
beta_ndvi(ndvi, model="linear") +
beta_urban(urban, model="linear") +
time(time, model="rw1") +
field(geometry, model = spde)
```

Likelihood:

```{r}
lik_st <- like(
family="binomial",
data = st_binom,
formula = n_pos ~ Intercept + beta_elev + beta_ndvi + beta_urban +
time + field,
Ntrials = n_tested
)
```

Fit:

```{r}
#| cache: true
fit_st_binom <- bru(components_st, lik_st)
summary(fit_st_binom)
```

## 12. Predict spatio-temporal prevalence (example: year = 2020)

```{r}
pred_grid$time <- 7

pred_st <- predict(
fit_st_binom,
newdata = pred_grid,
formula = ~ plogis(Intercept + beta_elev + beta_ndvi + beta_urban +
time + field),
n.samples = 500
)
```

Map for 2020:

```{r}
pred_st_r <- st_rasterize(pred_st["mean"], dx=10000, dy=10000)

ggplot() +
geom_stars(data = pred_st_r) +
scale_fill_viridis_c(name="Prev time 7") +
geom_sf(data=nga_admin1, fill=NA) +
theme_minimal()
```

# End of Day 1

By the end of today participants can:

-   Fit spatial binomial and Poisson models
-   Build meshes and PC priors
-   Predict over a grid
-   Generate exceedance probability maps
-   Fit basic spatio-temporal models using SPDE + RW1
