---
title: "Day 2 – Joint modelling and Non-stationary processes"
author: "Olatunji Johnson"
format: html
editor: visual
---


## Overview

Today we will cover two key advanced topics in spatial modelling:

1. **Joint modelling** of multiple malaria processes  
   - Shared spatial fields  
   - Process-specific fields  
   - Multiple outcomes (e.g., children vs pregnant women)

2. **Non-stationary spatial processes**  
   - Region-varying smoothness  
   - Covariate-driven nonstationarity  
   - Combining multiple spatial SPDEs

---

## Load Day 2 datasets

```{r}
library(tidyverse)

joint_malaria <- read_csv("data/joint_malaria_processes.csv")
nonstat_malaria <- read_csv("data/nonstationary_malaria_data.csv")

head(joint_malaria)
```

Quick visualisation

```{r}
library(ggplot2)

ggplot(joint_malaria, aes(utm_x, utm_y, colour = group)) +
geom_point() +
coord_equal() +
theme_minimal()
```

Quick visualisation (non-stationary dataset)

```{r}
ggplot(nonstat_malaria, aes(utm_x, utm_y, colour = prevalence_true)) +
geom_point() +
coord_equal() +
scale_colour_viridis_c() +
theme_minimal()
```



